Focasting The Trend of Vehicle Related Crime In Calgary In 2025

### 2024-11-28 paper/paper.qmd
Edit paper

library(tidyverse)
library(here)
library(knitr)

@@

Analyzing and projecting monthly crime counts for theft from vehicles and theft of vehicles in Calgary is the estimand of this study. I created a Bayesian regression model to assess the effects of time (year and month) and crime category on reported crime counts using historical data spanning January 2018 through October 2024. Offering a comprehensive knowledge of how vehicle-related crimes have developed and may continue to alter in the near future, the model catches both long-term patterns and seasonal changes. The research also offers category-specific forecasts, therefore stressing variations in the seasonal and temporal trends of theft from and of vehicles.

@@

# Display with kable
kable(column_df)
```

```{r}

Talk way more about it.

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

@@

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

@@

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

@@

Our results are summarized in @tbl-modelresults.

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

>>>>

library(tidyverse)
library(here)
library(knitr)
library(arrow)

cleaned_data <- read_parquet(here("data/02-analysis_data/calgary_crime.parquet"))
```

@@

Analyzing and projecting monthly crime counts for theft from vehicles and theft of vehicles in Calgary is the main aim of this study. I created a Bayesian regression model to assess the effects of time (year and month) and crime category on reported crime counts using historical data spanning January 2018 through October 2024. Offering a comprehensive knowledge of how vehicle-related crimes have developed and may continue to alter in the near future, the model catches both long-term patterns and seasonal changes. The research also offers category-specific forecasts, therefore stressing variations in the seasonal and temporal trends of theft from and of vehicles.

@@

# Display with kable
kable(column_df)
```
- Community: The dataset is spatially disaggregated into Calgary's various communities

Numeric Variable:

- Year: The year of each record indicate when the incidents were reported.

Categorical Variable:

- Category: Each entry is categorized into distinct crime types.

- month: The month of each incidents recorded. Range from 1-12 repeatedly each year

Response Variable:

- Crime Count: This variable quantifies the number of reported incidents of a specific crime type (e.g., theft from or of vehicles) within a given community and time period.

## Data Cleaning

Several steps was taken for better analysis of vehicle-related trend in Calgary. First, clean the names of all the columns. Second, the raw dataset contained granular records categorized by Community and specific details. These were aggregated to provide summarized counts of crimes (crime_count) by category, year, and month to align with analysis requirements. Third, a new time column was introduced in the cleaned data, combining year and month into a single date field (YYYY-MM-DD). This was essential for time-series analysis and visualization. Last, filter out the categories related to vehicle.

## Cleaned Data

The first 6 rows of the cleaned dataset are shown in @tbl-cleaned-data.

```{r}
#| label: tbl-cleaned-data
#| tbl-cap: Head of cleaned Calgary Crime data 
#| echo: false
#| warning: false

kable(head(cleaned_data))
```

## Response Variable

### Crime count of Theft FROM Vehicle

Summary statistics of Crime count of Theft FROM Vehicle are shown in the @tbl-response-theftfromvehicle

```{r}
#| label: tbl-response-theftfromvehicle
#| tbl-cap: Summary Statitic of crime_count(Theft FROM Vehicle)
#| echo: false
#| warning: false

kable(cleaned_data %>% filter(category == "Theft FROM Vehicle") %>% select(crime_count) %>% summary(crime_count))
```

The response variable, crime_count, represents the number of reported incidents of theft from vehicles within Calgary. The summary statistics reveal a significant variability in the crime counts, ranging from a minimum of 120 to a maximum of 1,619 reported cases. The median value is 916, indicating that half of the recorded months experienced fewer than 916 thefts, while the other half exceeded this value. The mean crime count, at 932.7, is slightly higher than the median, suggesting the presence of months with particularly high crime counts that may be skewing the average. The interquartile range (IQR), spanning from 788.2 (1st quartile) to 1,105.5 (3rd quartile), highlights the central tendency of crime occurrences, with most months falling within this range. These statistics emphasize the fluctuating nature of vehicle thefts in Calgary and provide a foundational understanding for modeling and forecasting future trends.

### Crime count of Theft OF Vehicle

Summary statistics of Crime count of Theft FROM Vehicle are shown in the @tbl-response-theftofvehicle

```{r}
#| label: tbl-response-theftofvehicle
#| tbl-cap: Summary Statitic of crime_count(Theft OF Vehicle)
#| echo: false
#| warning: false

kable(cleaned_data %>% filter(category == "Theft OF Vehicle") %>% select(crime_count) %>% summary(crime_count))
```

The response variable, crime_count, for Theft OF Vehicle incidents reflects the monthly reported counts of vehicles being stolen in Calgary. The dataset's summary statistics indicate that the crime counts range from a minimum of 146 to a maximum of 651 incidents. The median count is 424.5, suggesting that half of the months have fewer than 425 thefts, while the other half exceed this value. Interestingly, the mean crime count is 422.5, closely aligned with the median, suggesting a relatively symmetric distribution of the data. The interquartile range (IQR) spans from 361.2 (1st quartile) to 469.0 (3rd quartile), indicating that most monthly counts are concentrated within this range. These figures highlight the scale and regularity of vehicle theft in Calgary and form an essential basis for analyzing and forecasting trends in such crimes.

## Predictor Variable

### year

The variable year represents the timeline over which the crime data is recorded, providing insight into how vehicle-related crime counts have changed annually. By analyzing the data across different years, we can observe any long-term trends, such as consistent increases, decreases, or fluctuations in vehicle thefts. This variable helps us understand whether certain years experienced higher crime rates due to factors like economic conditions, population changes, or other external influences.

### month and category

The variables month and category provide a detailed view of how vehicle-related crimes are distributed across time and crime types. Month captures seasonal variations, showing how crime rates fluctuate throughout the year. For example, there might be a tendency for higher crime counts during certain seasons, such as summer or winter, influenced by weather conditions or holiday activities. Category distinguishes between "Theft FROM Vehicle" and "Theft OF Vehicle," allowing us to compare the frequency and patterns of these two crime types. Together, these variables highlight differences in the nature and timing of vehicle-related crimes, offering valuable insights into their temporal and categorical distributions.

## Trend of The Two Crime

The trend of Theft FROM Vehicle and Theft OF Vehicle are shown in @fig-combined-trend

```{r}
#| label: fig-combined-trend
#| tbl-cap: Trend of The Two Crime Combined
#| echo: false
#| warning: false
# Plot combined trends
ggplot(cleaned_data, aes(x = time, y = crime_count, color = category)) +
  geom_line() +
  labs(
    title = "Trend of Theft FROM Vehicle and Theft OF Vehicle",
    x = "Time",
    y = "Crime Count",
    color = "Category"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The trends for "Theft FROM Vehicle" and "Theft OF Vehicle" reveal distinct patterns over time. "Theft FROM Vehicle" consistently shows higher crime counts compared to "Theft OF Vehicle," indicating that it is a more prevalent issue. The trend for "Theft FROM Vehicle" demonstrates significant fluctuations, with peaks around 2019 and a general decline afterward, suggesting a reduction in these incidents in more recent years. In contrast, "Theft OF Vehicle" exhibits a relatively stable pattern with smaller variations over time. Although it remains less frequent than "Theft FROM Vehicle," its counts appear to decrease gradually, particularly after 2022. Both categories show declining trends toward the end of the observed period, which could reflect successful crime prevention measures or other external influences.

@@

The Bayesian regression model was developed to analyze and forecast the trends of vehicle-related crimes in Calgary, focusing on the categories "Theft FROM Vehicle" and "Theft OF Vehicle." The primary goal of this modeling is to understand the relationship between crime counts and key predictors such as time (year and month) and crime categories, as well as their interactions. By incorporating temporal and categorical variables, the model seeks to capture long-term trends, seasonal patterns, and differences between the two types of crimes. Ultimately, the model provides a foundation for projecting future crime rates, offering valuable insights that can support strategic decision-making and the development of targeted crime prevention initiatives for 2025 and beyond.

Background details and diagnostics are included in [Appendix -@sec-model-details].

@@

The Bayesian regression model is set up to examine the relationship between vehicle-related crime counts and the selected predictors. The response variable, $y_i$, represents the observed crime count for the i-th observation, where each observation corresponds to a specific month, year, and crime category. The predictors include a numeric variable for year, categorical variables for month and category, as well as interaction terms (category:month and category:year) to capture how crime patterns differ between the two crime types over time and across seasons.

The model assumes the form:

@@

Our results are summarized in

@@

```{r}
#| echo: false
#| eval: true

#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false
